### 2020-02-08 nlp_models.tex
updated main sections

\section{Comparison of NLP Models}
\label{sxn:nlp}

@@

\begin{figure}
    \centering
    \subfigure[Power law exponent ($\alpha$)]{
        \includegraphics[width=4.0cm]{img/GPT-alpha-hist.png}
        \label{fig:GPT-alpha-hist}
       }
    %\qquad
    \subfigure[Spectral Norm $(\log\Vert\mathbf{W}\Vert_{\infty})$]{
        \includegraphics[width=4.0cm]{img/GPT-snorm-hist.png}
        \label{fig:GPT-snorm-hist}
       }
   \caption{Comparison of PL exponents, $\alpha$, and log Spectral Norms, 
$(\log\Vert\mathbf{W}\Vert_{\infty})$, for the OpenAI GPT and GPT2 (small) pretrained models.}
\label{fig:GPT-hist}
\end{figure}

\paragraph{Scale Collapse.}
However, as seen in Figure \ref{fig:GPT-snorm-hist},
the ``bad'' GPT model has, spuriously, many smaller Spectral Norms $(\log\Vert\mathbf{W}\Vert_{\infty})$,
compared to the ``good'' GPT2--which violates the belief that smaller Spectral Norms are always better.
Indeed, because there are so many anonymously small $\Vert\mathbf{W}\Vert_{\infty}$,
it appears that the GPT model may be exhibiting a kind of \emph{scale collapse} and/or undesired
loss of Frobeinus mass.
This is an extremely important observation because it demonstrates that while the Spectral Norm
may correlate well with predicted test error, it is not a good indicator of the overall quality of a model,
and using it as an empirical metric may give spurious results when applied to poorly trained
or otherwise deficient models.

Note that Figure \ref{fig:GPT-snorm-hist} also shows some unusually large Spectral Norms, but,
here, simply correspond to the change in normalization of the word embedding layer(s) and can be ignored.
From Figure \ref{fig:gpt-snorm-layer} (below), we see that these correspond to the first embedding layer(s).
These layers appear to have a different effectvie normalization, and therefore a different scale.
For example, in GPT, most layers, the maximum eigenvalue $\lambda_{max}\sim\mathcal{O}(10-100)$,
but in the first embedding layer, the maximum is of order N (the number of words in the embedding), or
 $\lambda_{max}\sim\mathcal{O}(10^{5})$.  For GPT and GPT2, we treat all layers as-is (although one may to normalize
the first 2 layers by  $\mathbf{X}$ by $\frac{1}{N}$, or to treat them as an outlier).
Here, we do not include them in our computed average metrics in Table \ref{table:nlp},
and do not include them in the histogram plot in Figure \ref{fig:GPT-snorm-hist}.

@@

\begin{figure}[H]
    \centering
    \subfigure[Power Law exponent $(\alpha)$]{
        \includegraphics[width=4.0cm]{img/GPT-alpha-depth.png}
        \label{fig:gpt-alpha-layer}
    }
    \subfigure[Spectral Norm $(\log\Vert\mathbf{W}\Vert_{\infty})$]{
        \includegraphics[width=4.0cm]{img/GPT-snorm-depth.png}
        \label{fig:gpt-snorm-layer}
    }
    \caption{Comparison of Correlation Flow and Spectral Norm for OpenAI GPT and GPT2   }
    \label{fig:gpt-alpha-layers}
\end{figure}

@@

\begin{figure}[t]
    \centering

    \subfigure[PL exponent ($\alpha$)]{
        \includegraphics[width=4.0cm]{img/GPT2*_fnl_alpha_hist.png}
        \label{fig:gpt2-alpha-hist}
    }
    %\qquad
    %%\subfigure[log Spectral Norm]{
    %    \includegraphics[width=4.0cm]{img/GPT2*_fnl_spectralnorm_hist.png}
    %    \label{fig:gpt2-snorm-hist}
   % }
    %\qquad
    \subfigure[log Alpha Norm]{
        \includegraphics[width=4.0cm]{img/GPT2*_fnl_logpnorm_hist.png}
        \label{fig:gpt2-pnorm-hist}
    }
    \caption{PL exponents ($\alpha$), log Spectral norm $(\log\Vert\mathbf{W}\Vert_{\infty})$, and log Alpha norm $(\log\Vert\mathbf{X}\Vert_{\alpha}^{\alpha})$ for different size models in the GPT2 architecture series.  (Plots omit the first 2 (embedding) layers, because they are normalized differently giving anamolously large values.)}
    \label{fig:gpt2-histograms}
\end{figure}

?

>>>>

section{Comparison of NLP Models}
\label{sxn:nlp}

@@

\paragraph{Scale Collapse in Poorly Trained Models}
However, as seen in Figure \ref{fig:GPT-snorm-hist},
the ``bad'' GPT model has, spuriously, many smaller Spectral Norms $(\log\Vert\mathbf{W}\Vert_{\infty})$,
compared to the ``good'' GPT2--which violates the belief that smaller Spectral Norms are always better.
Indeed, because there are so many anonymously small $\Vert\mathbf{W}\Vert_{\infty}$,
it appears that the GPT model may be exhibiting the \emph{scale collapse} observed
in the distilled CV models (above).
This is an extremely important observation because it demonstrates that while the Spectral Norm
may correlate well with predicted test error, it is not a good indicator of the overall quality of a model,
and using it as an empirical metric may give spurious results when applied to poorly trained
or otherwise deficient models.

\begin{figure}[h]
    \centering
    \subfigure[Power law exponent ($\alpha$)]{
        \includegraphics[width=4.0cm]{img/GPT-alpha-hist.png}
        \label{fig:GPT-alpha-hist}
       }
    %\qquad
    \subfigure[Spectral Norm $(\log\Vert\mathbf{W}\Vert_{\infty})$]{
        \includegraphics[width=4.0cm]{img/GPT-snorm-hist.png}
        \label{fig:GPT-snorm-hist}
       }
   \caption{Comparison of PL exponents, $\alpha$, and log Spectral Norms, 
$(\log\Vert\mathbf{W}\Vert_{\infty})$, for the OpenAI GPT and GPT2 (small) pretrained models.}

\label{fig:GPT-hist}
\end{figure}

Note that Figure \ref{fig:GPT-snorm-hist} also shows some unusually large Spectral Norms, but,
here, simply correspond to the change in normalization of the word embedding layer(s) and can be ignored.
From Figure \ref{fig:gpt-snorm-layer} (below), we see that these correspond to the first embedding layer(s).
These layers appear to have a different effective normalization, and therefore a different scale.
\charles{We discuss this further in the Appendix}
%For example, in GPT, most layers, the maximum eigenvalue $\lambda_{max}\sim\mathcal{O}(10-100)$,
%but in the first embedding layer, the maximum is of order N (the number of words in the embedding), or
% $\lambda_{max}\sim\mathcal{O}(10^{5})$.  For GPT and GPT2, we treat all layers as-is (although one may to normalize
%the first 2 layers by  $\mathbf{X}$ by $\frac{1}{N}$, or to treat them as an outlier).
Here, we do not include them in our computed average metrics in Table \ref{table:nlp},
and do not include them in the histogram plot in Figure \ref{fig:GPT-snorm-hist}.

@@

\begin{figure}[htb]
    \centering
    \subfigure[Power Law exponent $(\alpha)$]{
        \includegraphics[width=4.0cm]{img/GPT-alpha-depth.png}
        \label{fig:gpt-alpha-layer}
    }
    \subfigure[Spectral Norm $(\log\Vert\mathbf{W}\Vert_{\infty})$]{
        \includegraphics[width=4.0cm]{img/GPT-snorm-depth.png}
        \label{fig:gpt-snorm-layer}
    }
    \caption{Comparison of Correlation Flow and Spectral Norm for OpenAI GPT and GPT2   }
    \label{fig:gpt-alpha-layers}
\end{figure}

@@

\begin{figure}[htb]
    \centering

    \subfigure[PL exponent ($\alpha$)]{
        \includegraphics[width=4.0cm]{img/GPT2*_fnl_alpha_hist.png}
        \label{fig:gpt2-alpha-hist}
    }
    \subfigure[log Alpha Norm]{
        \includegraphics[width=4.0cm]{img/GPT2*_fnl_logpnorm_hist.png}
        \label{fig:gpt2-pnorm-hist}
    }
    \caption{PL exponents ($\alpha$), log Spectral norm $(\log\Vert\mathbf{W}\Vert_{\infty})$, and log Alpha norm $(\log\Vert\mathbf{X}\Vert_{\alpha}^{\alpha})$ for different size models in the GPT2 architecture series.  (Plots omit the first 2 (embedding) layers, because they are normalized differently giving anamolously large values.)}
    \label{fig:gpt2-histograms}
\end{figure}
